---
title: "MAMBO vignette"
output: html_document
date: "2025-03-24"
---

## R Markdown

This is a tutorial for running metabarcoding data sets through MAMBO (Metabarcoding Analysis of Modeled Bayesian Occurrences). To run MAMBO, you will need two sets each of 1) ASV tables and 2) taxonomy tables. You may also include an optional metadata file.

The ASV tables should have samples as columns, ASVs as rows, and an empty A1 cell. The cell values should contain raw (untransformed) read counts.

The taxonomy tables should have ASVs as rows and taxonomy ranks in subsequent columns, which each rank in its own column.

```{r libraries}
library(mambo)
library(sprex)
library(dplyr)
library(ggplot2)
library(here)
```

#### Import your ASV tables
```{r ASV data}
asv.16s <- read.csv(file=here::here("Data", "ASV tables", 
                              "Merged2018_16S_otu_filtered.csv"),
                    row.names=1)

asv.18s <- read.csv(file=here::here("Data", "ASV tables",
                              "Merged2018_18S_otu_filtered.csv"),
                    row.names=1)
```

#### Import your taxonomy tables
```{r tax data}
tax.16S <- read.csv(file=here("Data", "Taxonomy tables", 
                              "Merged2018_16S_taxa_filtered.csv"))

tax.18s <- read.csv(file=here("Data", "Taxonomy tables", 
                              "Merged2018_18S_taxa_filtered.csv"))
```

#### Convert the ASV tables to dataframes and check some basic features of your data

```{r features per sample}
asv.16s.counts <- data.frame(Sample = names(asv.16s), 
                             ASVs = colSums(asv.16s != 0))
rownames(asv.16s.counts) <- NULL

hist <- ggplot(asv.16s.counts, aes(x=ASVs)) +
  geom_histogram(aes(y=..density..), 
                 position="identity", binwidth=5)

# samples per feature
asv.16s.T <- setNames(data.frame(t(asv.16s)),
                      rownames(asv.16s))
asv.16s.samps <- data.frame(ASV = names(asv.16s.T), 
                             Samples = colSums(asv.16s.T != 0))
rownames(asv.16s.samps) <- NULL

hist <- ggplot(asv.16s.samps, aes(x=Samples)) +
  geom_histogram(aes(y=..density..), 
                 position="identity", binwidth=5)

```

#### Run MAMBO and check the labels
```{r mambo}

result <- mambo(
  resp.label = '18s',
  resp.counts = asv.18s,
  pred.label = '16s',
  pred.counts = asv.16s,
  nrep=10,
  chains=3,
  adapt=100,
  burnin=1000,
  total.samples=1000,
  thin=1,
  run.label="mambo",
  output.log=TRUE
)

result$labels
```

#### To skip the modeling and just run the PCAs, set Bayesian=FALSE
```{r mambo}

result <- mambo(
  resp.label = '18s',
  resp.counts = asv.18s,
  pred.label = '16s',
  pred.counts = asv.16s,
  nrep=10,
  chains=3,
  adapt=100,
  burnin=1000,
  total.samples=1000,
  thin=1,
  run.label="mambo",
  bayesian=FALSE,
  output.log=TRUE
)

result$labels
```

#### Check the results of the first replicate

```{r results}

result$reps[[1]]$pca$`16S`
```

### Plot the PCAs for each marker gene ("locus")
```{r PCA}

# default
p16s <- plotPCs(result, "16S", pc.x=1, pc.y=2, type="ellipse")

# with confidence ellipses
p16s <- plotPCs(result, "16S", pc.x=1, pc.y=2, type="density")
```

```{r PCA}

# default
p18s <- plotPCs(result, "18S", pc.x=1, pc.y=2, type="ellipse")

# with confidence ellipses
p18s <- plotPCs(result, "18S", pc.x=1, pc.y=2, type="density")
```

#### Check which ASVs are outliers, ie driving the separations
```{r outliers}

outliers.16s.pc1 <- outlierLoadings(result, 
                                    locus = '16S', 
                                    pc = 1)

outliers.18s.pc1 <- outlierLoadings(result, 
                                    '18S', 
                                    pc = 1)

head(outliers.16s.pc1)
```

#### Plot the summary of sign switching
```{r signs}

mambo:::switchSummary(result)
```

### Extract PC scores
```{r scores}

scores.16s <- extractPCA(result)$scores[['16S']]
scores.18s <- extractPCA(result)$scores[['18S']]
```

#### Identify taxonomy of outlier ASVs
```{r tax}

# Add column name to match ASV tibble column name if it doesn't already
colnames(tax.16S)[1] <- "asv"
colnames(tax.18s)[1] <- "asv"

# Merge the tibbles
asvs_out_16S_tax <- left_join(outliers.16s.pc1, tax.16S,
                              by="asv")
asvs_out_18S_tax <- left_join(outliers.18s.pc1, tax.18s, 
                              by="asv")

head(asvs_out_16S_tax)
```

```{r save data}
write.csv(asvs_out_16S_tax, "CCE_16S_outlier_tax.csv",
          row.names=FALSE)

write.csv(asvs_out_18S_tax, "CCE_18S_outlier_tax.csv",
          row.names=FALSE)
```

#### If you have an existing MAMBO output saved as a .rds file, you can import it and extract the existing data

```{r import rds}
file = here("OneDrive - SCCWRP", "Bioinformatics", "MAMBO", "GOMECC", "outputs", "mambo.20240917_193658.rds")
gomecc = readRDS(file)
```