---
title: "MAMBO Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "",
  prompt = FALSE,
  fig.width = 8,
  fig.height = 8
)
```

This is a tutorial for running metabarcoding data sets through MAMBO (Metabarcoding Analysis of Modeled Bayesian Occurrences). You should start with two ASV tables, each of which has columns as samples and rows as ASVs.

```{r libraries}
library(mambo)
```

## ASV Tables

ASV count tables should be loaded as R data frames or matrices. The values in the tables represent the occurrence (number of reads) for each ASV (rows) in each sample (columns). The table should have both row and column names uniquely identifying ASVs and samples.

We supply two sets of two example ASV count tables in `mambo`: 16S and 18S data from the California Current Ecosystem ('`counts.16s`' and '`counts.18s`'), and 12S and COI data from Blackman ('`counts.12s`' and '`counts.COI`').
```{r data}
counts.16s[1:8, 1:4]
```

We also supply metadata for the CCE samples as `sample.meta`:

```{r}
str(sample.meta)
```


## Filter data

First we'll extract the Lasker samples from the CCE 16S and 18S data and filter them using the default values. The `filterCountData()` function will report the number of ASVs and samples removed at each step.

```{r}
lasker.samples <- sample.meta$sample[sample.meta$cruise == 'Lasker']

lasker.16s <- filterCountData(counts.16s[, lasker.samples])
lasker.18s <- filterCountData(counts.18s[, lasker.samples])
```


## Run MAMBO 

To run the main function, `mambo()`, we need to supply labels for the response and predictor datasets and the response and predictor count tables, along with the number of replicates (`nrep`) and parameters for the Bayesian MCMC run (see `?mambo`). One can also specify a label for the run ('`run.label`') that will be used for the log and results file. By default, this is set to '`mambo`'. 
```{r, echo = TRUE, cache=TRUE}
results <- mambo(
  resp.counts = lasker.18s, 
  resp.label = 'Lasker_18S', 
  pred.counts = lasker.16s, 
  pred.label = 'Lasker_16S', 
  nrep = 10, 
  chains = 3,
  adapt = 100,
  burnin = 1000,
  run.label = 'Lasker'
)
```

To skip the Bayesian modeling and just compute diversity and run the PCAs, set `bayesian = FALSE`:
```{r, echo = TRUE, eval = FALSE, cache=TRUE}
results <- mambo(
  resp.counts = lasker.18s, 
  resp.label = 'Lasker_18S', 
  pred.counts = lasker.16s, 
  pred.label = 'Lasker_16S', 
  nrep = 10, 
  bayesian = FALSE,
  label = 'Lasker'
)
```

While `mambo()` is running, it will output progress on the screen and to a log file named, '<label>.YYYYMMDD_HHmmSS.log'. When the run is complete, it will write its results to an .RDS file in the working directory that is named the same as the log file, but ends in '.rds'. This file can be read into an R object using `readRDS()`:

```{r, eval=FALSE}
results <- readRDS('Lasker.Lasker.20260128_132951.rds')
```

## Examine model parameters and outputs

The output of `mambo()` is a list with five elements:

1) The filename of the .RDS file that was written:  
```{r}
results$filename
```

2) The labels that were provided to the function for each locus:
```{r}
results$labels
```

3) The MCMC parameters:
```{r}
results$params
```

4) The run times for the model:
```{r}
results$run.time
```

5) A list of the results for every replicate (`$reps`). 
```{r}
length(results$reps)
```

Each element in the `$reps` list contains results for diversity and PCAs for each locus, and the Bayesian modelling for every replicate: 
```{r}
names(results$reps[[1]])
```

The `$total.diversity` element is a vector giving the total ASV diversity for each locus:
```{r}
results$reps[[1]]$total.diversity
```

The `$sample.diversity` element is a data.frame giving the ASV diversity for each sample (rows) at each locus (columns):
```{r}
head(results$reps[[1]]$sample.diversity)
```

The `$pca` element contains PCA results for each locus, where `$rotation` has the loading values of ASVs on the PCs, `$x` has the scores for each sample on each PC, and `$importance` gives summary information on variance explained by each PC:
```{r}
str(results$reps[[1]]$pca[['Lasker_16S']])
```

However, the function `extractPCA()`, will return this information for all replicates in a way that is easier to manipulate. It returns a list with one element for loadings, scores, and importance, where each is a data frame:
```{r}
pca <- extractPCA(results)
str(pca)
```


If a Bayesian analysis was conducted, then the `$post.smry` element will be a matrix providing a summary of the posteriors from the Bayesian model for each parameter:
```{r}
head(results$reps[[1]]$post.smry)
```

... and `$post.list` will contain the actual posteior sample:
```{r}
str(results$reps[[1]]$post.list)
```

If no Bayesian model was conducted, both of these will be `NULL`.



## Plot results

### Plot ASV diversity (the effective number of ASVs).

```{r}
plotDiversity(
  results, 
  sample.df = sample.meta,
  ellipse.fill = 'depth_m'
)
```

This demonstrates how 18S diversity (y-axis) changes relative to 16S diversity (x-axis). The red line shows a 1:1 correspondence and ellipses show the 95-th percentile of the replicates in each axis. We have color-coded each ellipse relative to the depth that the sample was taken from to highlight a general pattern of increasing diversity with depth.

### Plot distribution of percent variance accounted for by each Principal Component

```{r}
screePlot(results)
```


### Biplot of selected Principal Components for a locus

```{r}
plotPCs(
  results, 
  locus.x = 'Lasker_16S', 
  pc.x = 1, 
  pc.y = 2, 
  sample.df = sample.meta, 
  ellipse.fill = 'depth_m'
)
```

```{r}
plotPCs(
  results, 
  locus.x = 'Lasker_18S', 
  pc.x = 2, 
  pc.y = 3, 
  sample.df = sample.meta, 
  ellipse.fill = 'depth_m'
)
```

## Summarize the posterior of predictive switches 
Here we summarize the first four predictor and response PCs:
```{r switch, message=FALSE, warning=FALSE}
sw.smry <- switchSummary(results, min.p = 0.9, resp.pcs = 1:4, pred.pcs = 1:4)
```
 
The `$smry` element is a data frame which identifies relationships between response and predictor PCS that were identified as significant based on the Bayesian model:

```{r}
sw.smry$smry
```

We can plot those relationships using the `plotPCs()` function and supplying both the `x` and `y` loci and their respective PC numbers:
```{r}
plotPCs(
  results, 
  locus.x = 'Lasker_16S', 
  locus.y = 'Lasker_18S', 
  pc.x = 1, 
  pc.y = 1, 
  sample.df = sample.meta, 
  ellipse.fill = 'depth_m'
)
```

```{r}
plotPCs(
  results, 
  locus.x = 'Lasker_16S', 
  locus.y = 'Lasker_18S', 
  pc.x = 2, 
  pc.y = 2, 
  sample.df = sample.meta, 
  ellipse.fill = 'depth_m'
)
```

 
## Identify ASVs that have unusually high or low loadings. 
Contrasts between these 'outlier' ASVs define the source of variation for that PC.
```{r outliers_16S, paged.print=FALSE}
outliers.16s.pc1 <- outlierLoadings(results, locus = 'Lasker_16S', pc = 1)

outliers.16s.pc1
```

## Identify taxonomy of outlier ASVs

```{r taxa}
# Add column name to match ASV tibble column name if it doesn't already
taxa.16s$asv <- rownames(taxa.16s)

# Merge the tibbles
asvs_out_16s_tax <- dplyr::left_join(outliers.16s.pc1, taxa.16s, by = "asv")

head(asvs_out_16s_tax)
```