# Try running mambo
# KP 01/24/24

```{r}
library(devtools)
```

```{r}
load_all("/Users/kpitz/github/nvpatin/MAMBO/mambo")

```

# from vignette
```{r libraries}
# library(mambo)
library(tidyverse)
```

#### Import your ASV tables
```{r}
# use data here:
# "/Users/kpitz/github/nvpatin/MAMBO/Data/merged_data/no_duplicates/"

df_16s <- read.csv("/Users/kpitz/github/nvpatin/MAMBO/Data/merged_data/no_duplicates/Merged2018_16S_otu_filtered.csv", row.names=1)

df_18s <- read.csv("/Users/kpitz/github/nvpatin/MAMBO/Data/merged_data/no_duplicates/Merged2018_18S_otu_filtered.csv", row.names=1)
```

#### Run MAMBO 

mambo <- function(
    resp.label, resp.counts, 
    pred.label, pred.counts,
    nrep = 10,
    chains = 3,
    adapt = 100,
    burnin = 1000,
    total.samples = 1000,
    thin = 1,
    run.label = 'mambo',
    output.log = TRUE
)

```{r}
test_Results <- mambo('18S', df_18s, '16S', df_16s)
```

```{r}
#test_Results2 <- mambo('16S', df_16s, '18S', df_18s)
```

## Plot Results


```{r}
plotPCs(test_Results, '16S')
```
```{r}
plotPCs(test_Results, '18S')
```



# Try out individual PCA:
### Functions
ranPCA <- function(beta.params) {
  prob <- t(ranRelPct(beta.params))
  log(prob / (1 - prob)) |>
    stats::prcomp() |>
    summary()
}
ranRelPct <- function(beta.params) {
  # draw random sample from beta distribution with shape parameters 'p'
  pct <- apply(beta.params, c(1, 2), function(p) stats::rbeta(1, p[1], p[2]))

  # normalize random percents to unity and return matrix
  # (transposed back to original dimensions)
  pct <- t(pct / rowSums(pct))
  rownames(pct) <- dimnames(beta.params)[[2]]
  colnames(pct) <- dimnames(beta.params)[[1]]
  pct
}
betaParams <- function(x) {
  # read data if filename is given
  if(is.character(x)) x <- utils::read.delim(x, row.names = 1)
  x <- as.matrix(x)

  if(length(rownames(x)) == 0) stop('count tables need ASV names for the rows')
  if(length(colnames(x)) == 0) stop('count tables need sample names for the columns')
  
  # by-sample (columns) coverage
  coverage <- colSums(x)

  # fit beta shape parameters (transpose matrix for recycling of coverage vector)
  tx <- t(x)
  result <- array(c(tx + 1, coverage - tx + 1), dim = c(dim(tx), 2))
  dimnames(result) <- list(colnames(x), rownames(x), c('shape1', 'shape2'))
  
  result
}
```{r}
# Try out single PC for 16S
test_betaParams <- betaParams(df_16s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
```

```{r}
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID")
df %>%
  ggplot(aes(x=PC1, y=PC2)) +
  geom_point()
# df %>%
#   ggplot(aes(x=PC2, y=PC3)) +
#   geom_point()
```

```{r}
# Try out single PC for 18S
test_betaParams <- betaParams(df_18s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
```

```{r}
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID")
df %>%
  ggplot(aes(x=PC1, y=PC2)) +
  geom_point()
# df %>%
#   ggplot(aes(x=PC2, y=PC3)) +
#   geom_point()
```



# How variable are 16S results naturally? Or is this a sign flipping problem?

```{r}
# Try out single PC for 16S
test_betaParams <- betaParams(df_16s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID") %>%
  mutate(cruise = case_when(str_detect(sampleID, "CN18S") ~ "CN18S",
                            str_detect(sampleID, "CN18F") ~ "CN18F",
                            str_detect(sampleID, "Lasker") ~ "Lasker",
                            TRUE ~ "unknown")) %>%
  select(sampleID, PC1, PC2,PC3, cruise)
df %>%
  ggplot(aes(x=PC1, y=PC2, color=cruise)) +
  geom_point() #+
  #geom_text(aes(label=sampleID))
# df %>%
#   ggplot(aes(x=PC2, y=PC3, color=cruise)) +
#   geom_point()
```

```{r}
# Try out single PC for 16S
test_betaParams <- betaParams(df_16s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID") %>%
  mutate(cruise = case_when(str_detect(sampleID, "CN18S") ~ "CN18S",
                            str_detect(sampleID, "CN18F") ~ "CN18F",
                            str_detect(sampleID, "Lasker") ~ "Lasker",
                            TRUE ~ "unknown")) %>%
  select(sampleID, PC1, PC2,PC3, cruise)
df %>%
  ggplot(aes(x=PC1, y=PC2, color=cruise)) +
  geom_point() #+
  #geom_text(aes(label=sampleID))
# df %>%
#   ggplot(aes(x=PC2, y=PC3, color=cruise)) +
#   geom_point()
```
```{r}
# Try out single PC for 16S
test_betaParams <- betaParams(df_16s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID") %>%
  mutate(cruise = case_when(str_detect(sampleID, "CN18S") ~ "CN18S",
                            str_detect(sampleID, "CN18F") ~ "CN18F",
                            str_detect(sampleID, "Lasker") ~ "Lasker",
                            TRUE ~ "unknown")) %>%
  select(sampleID, PC1, PC2,PC3, cruise)
df %>%
  ggplot(aes(x=PC1, y=PC2, color=cruise)) +
  geom_point() #+
  #geom_text(aes(label=sampleID))
# df %>%
#   ggplot(aes(x=PC2, y=PC3, color=cruise)) +
#   geom_point()
```


```{r}
# Try out single PC for 16S
test_betaParams <- betaParams(df_16s)
test_ranPCR <-  ranPCA(test_betaParams)
test_ranRelPct <- ranRelPct(test_betaParams)
# Plot
df <- as_tibble(test_ranPCR$x, rownames = "sampleID") %>%
  mutate(cruise = case_when(str_detect(sampleID, "CN18S") ~ "CN18S",
                            str_detect(sampleID, "CN18F") ~ "CN18F",
                            str_detect(sampleID, "Lasker") ~ "Lasker",
                            TRUE ~ "unknown")) %>%
  select(sampleID, PC1, PC2,PC3, cruise)
df %>%
  ggplot(aes(x=PC1, y=PC2, color=cruise)) +
  geom_point()
# df %>%
#   ggplot(aes(x=PC2, y=PC3, color=cruise)) +
#   geom_point()
```
# Is sign flipping of PC2 causing this 16S results?

### functions
plotPCs <- function(results, locus, pc.x = 1, pc.y = 2, ellipse.p = 0.95) {
  scores <- extractPCA(results)$scores[[locus]]
  
  purrr::imap(split(scores, scores$sample), function(df, i) {
    x <- dplyr::filter(df, pc == pc.x)$score
    y <- dplyr::filter(df, pc == pc.y)$score
    car::dataEllipse(x, y, levels = ellipse.p, draw = FALSE) |> 
      as.data.frame() |> 
      dplyr::mutate(sample = i)
  }) |> 
    dplyr::bind_rows() |> 
    ggplot2::ggplot() +
    ggplot2::geom_hline(yintercept = 0, color = 'darkred') +
    ggplot2::geom_vline(xintercept = 0, color = 'darkred') +
    ggplot2::geom_polygon(
      ggplot2::aes(x, y, group = sample),
      fill = NA,
      color = 'black'
    ) +
    ggplot2::labs(
      x = paste0('PC', pc.x),
      y = paste0('PC', pc.y)
    ) +
    ggplot2::theme_minimal()
}

extractPCA <- function(results) {
  pca.list <- pcaList(results)
  
  loadings <- lapply(pca.list, function(x) {
    res <- abind::abind(x$rotation, along = 3) |> 
      apply(2, function(x) {
        to.switch <- sign(x[1, ]) != sign(x[1, 1])
        x[, to.switch] <- x[, to.switch] * -1
        x
      }, simplify = FALSE) |> 
      abind::abind(along = 3) |> 
      aperm(c(1, 3, 2))
    dimnames(res)[[3]] <- 1:dim(res)[3]
    as.data.frame.table(res) |> 
      stats::setNames(c('asv', 'pc', 'rep', 'loading')) |> 
      dplyr::mutate(
        asv = as.character(asv),
        pc = as.numeric(gsub('PC', '', pc)),
        rep = as.numeric(rep)
      )
  })
  
  scores <- lapply(pca.list, function(x) {
    res <- abind::abind(x$x, along = 3) |> 
      apply(2, function(x) {
        to.switch <- sign(x[1, ]) != sign(x[1, 1])
        x[, to.switch] <- x[, to.switch] * -1
        x
      }, simplify = FALSE) |> 
      abind::abind(along = 3) |> 
      aperm(c(1, 3, 2))
    dimnames(res)[[3]] <- 1:dim(res)[3]
    as.data.frame.table(res) |> 
      stats::setNames(c('sample', 'pc', 'rep', 'score')) |> 
      dplyr::mutate(
        sample = as.character(sample),
        pc = as.numeric(gsub('PC', '', pc)),
        rep = as.numeric(rep)
      )
  })
  
  list(loadings = loadings, scores = scores)
}

```{r}
# Extract PCA results from mambo run
test_extractPCA <- extractPCA(test_Results)
```

# Look at range of values for same sample within a PC (16S)

```{r}
library(magrittr)
df <- as_tibble(test_extractPCA$scores[[2]])
# get stats by unique sample
PC_stats <- df %>% 
  filter(pc<4) %>%
  #group_by(pc) %>%
  group_by(pc, sample) %>%
  mutate(num_reps = sum(rep)) %>%
  mutate(avg_score = mean(score)) %>%
  mutate(min_score = min(score)) %>%
  mutate(max_score = max(score)) %>%
  distinct(sample, .keep_all = TRUE) %>%
  select(pc,sample, num_reps, avg_score, min_score, max_score) %>%
  mutate(range = max_score - min_score) %>%
  ungroup()

PC_stats %>%
  ggplot(aes(x=pc, y=range)) +
  geom_point()

```
# Look at range of values for same sample within a PC (18S)

```{r}
library(magrittr)
df <- as_tibble(test_extractPCA$scores[[1]])
# get stats by unique sample
PC_stats <- df %>% 
  filter(pc<4) %>%
  #group_by(pc) %>%
  group_by(pc, sample) %>%
  mutate(num_reps = sum(rep)) %>%
  mutate(avg_score = mean(score)) %>%
  mutate(min_score = min(score)) %>%
  mutate(max_score = max(score)) %>%
  distinct(sample, .keep_all = TRUE) %>%
  select(pc,sample, num_reps, avg_score, min_score, max_score) %>%
  mutate(range = max_score - min_score) %>%
  ungroup()

PC_stats %>%
  ggplot(aes(x=pc, y=range)) +
  geom_point()

```




